Semi-quantitative petrographic characterization of Bronze Age quartzite tools – an exploratory approach

### 2022-09-12 analysis/paper/paper.Rmd
Updated text and structure

Petrographic characterization and provenance studies are instrumental in the modelling and inference of trade and exchange networks among prehistoric societies (REF). The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery. Although the petrographic characterization of the material mainly establishes a point of geological origin for the object it is still an important first step in reconstructing the *chaîne opératoire*. Flint and obsidian have both been central to the study of trade and exhange networks in European prehistory, due in part to the well-surveyed geological sources for these. The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. Prieto et al. [-@prieto_etal19 p. 15] highlights the current bias that exists in the understanding of raw material management in regions where local flint and obsidian sources are non-existent^[This is not to say that flint was not available or used in these regions]. In northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region. While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. While models of raw material sourcing and trade and exchange of slate and flint within Northern Sweden has been put forth [@halen94 pp. 115-122; @lundberg97 pp. 161-171], there has been little development of on the topic of quartz and quartzite. A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the modern day landscape. The tentative interpretations of lithic material sourcing within northern Sweden tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12]. This does not, however, account for the possibility that the material was sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250]. Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.

Prieto et al. [-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature, and different researchers have thus emphasized different features as being characteristic of the material [@howard05]. Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies. In the case of archaeological research it is worth considering situations where destructive techniques simply are not applicable. Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution. While destructive methods may prohibit access to such material there are still non-destructive alternatives that can generate valuable information. Aside from being non-destructive spectroscopic instrumentations feature fast measurements, making it an attractive option for field-based screening purposes. While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other instrumentations that may provide a deeper insight into the structural properties. The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. The same can be said for Raman spectroscopy, which has been applied extensively in the Cultural Heritage sector and in the analysis of artwork and manuscripts [@vandenabeele12].

@@

The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from excavated and surveyed sites, including stray finds, in Västerbotten County, Sweden. The sites are distributed along the Ångermanälven river valley, from the lake Kultsjön in the north, down towards the town of Åsele in the south. A dataset of `r length(tot_uniq_samples)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF. The material features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint and some being almost fully translucent. The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent).

@@

The dataset includes three main groups of registered petrographic material: quartz, quartzite, and brecciated quartz (Swe. "*brecciekvarts*"). Those that had no prior classification available were sorted into either the quartz or quartzite category based on visual inspection. This resulted in: `r length(bquartz)` brecciated quartz samples, `r length(quartz)` quartz samples, and `r length(quartzite)` quartzite samples. The inclusion of quartz and brecciated quartz in the study serves two purposes. The first being to test the potential for separating quartz from quartzitebased on their chemical and structural characteristic, and the second related purpose is to account for potential mistakes in the original classification. When analysing the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67 p.8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden. Åhman describes the material as easily mistaken for an unusually pure quartzite, however, the fragmented configuration of quartz can be identified via microscopic analysis. *Brecciated quartz* in particular became a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g. @holm91 p.24]. As the material frequently has been characterized based on megascopic characteristics, despite the central features necessary for characterization being microscopic, there is value in including a test group in the study.

@@

Pre-processing of the data is different for each instrumentation and based on relevant literature.

@@

NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe. The ASD features a detection range of 350 nm – 2 500 nm, with spectral sampling (resp. spectral resolution) of 1.4 nm (resp. 3 nm) in the visible and near infrared range and 1.1 nm (resp. 10 nm) in the short‐wave infrared range. Each object was measured against a white reference background and the collected measurements were averaged for each artefact. The visible spectrum (~ 400 nm - 750 nm) and the NIR range (~ 1 000 nm – 2 500 nm) were analysed separately, with the former focused on colour analysis and the latter the chemical structure of the material. The data was mean-centred before PCA modelling.

@@

The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling. As the resulting dataset also features values below the Limit of Detection (LOD) these were treated as NULL (or "missing values") during PCA. A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement. Since missing values are not permitted during PCA these had to be imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data. Two methods for imputation can be implemented, iterative or regularized, and since the former suffers when there is a large number of missing values the latter was used.

Multivariate analysis methods, such as PCA, are also useful for overcoming some of the issues related to so-called "data closure" [@reimann_etal12]. The compositional data add up to a constant (100%), which is described as the *closed sum* effect. Correlation analysis is directly affected by this *closure* effect and therefore not suitable. With suitable preprocessing the PCA model can reveal the structure and relations between the variables (elements) more accurately.

@@

## O2PLS

@@

O2PLS was implemented in R [@rcoreteam21] using the *OmicsPLS* package [@bouhaddani_etal18]. The data was pre-processed according to the above descriptions, and then cross-validation was used in order to determine the number of components to fit to the model.

\pagebreak

@@

```{r nir-savitzky-transform-alt, fig.cap= "nir savitzky comparison", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform-alt.R"), echo = FALSE}
quietly(source("./analysis/paper/003-nir-savitzky-transform-alt.R", encoding = "UTF-8"))
knitr::include_graphics("../figures/003-nir-savitzky-transform-alt.png", error = FALSE)
```

@@

Based on the evaluation of the spectra a couple of significant peaks can be identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm. These features are all the results of overtones and vibration combinations of hydroxyl (OH) groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water (fig. \@ref(fig:nir-peak-summary)) [@hunt77 p. 508-510; @sciuto_etal19]. OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occuring in the 2 000 nm - 2 500 nm region. As a general rule, the 2 200 nm and 2 350 nm peak positions have been related to aluminium (Al) and magnesium (Mg) combinations [@hunt77 p.509; @clark_etal90 p.12664]. While there are exceptions, a more intense 2 200 nm band has typically been related to AlOH, and a more intense 2 350 nm band to MgOH. In the current spectra the 2 200 nm band is typically the more intense band.

```{r nir-peak-summary, fig.cap= "nir peak summary", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/002-nir-peak-summary.R"), echo = FALSE}
quietly(source("./analysis/paper/002-nir-peak-summary.R", encoding = "UTF-8"))
knitr::include_graphics('../figures/002-nir-peak-summary.png', error = FALSE)
```

@@

While the explained variance of PC 3 is low, (`r nir.pc3var` %), it still contains a number of the aforementioned features visible in the spectra (inverse to PC 1). Adding PC 3 to a score plot (fig. \@ref(fig:nir-pca-fig):B), however,  there seems to be little new information gained apart from a couple of samples, mainly from the light group, that break away from the remaining samples.

@@

The resulting PCA model (fig. \@ref(fig:raman-pca)) shows a clear trend along PC 1 (`r raman.pc1var` %) related to the quartz and D - G peaks. Samples where the D/G peaks dominate in terms of intensity are clustered towards the positive side of PC 1, with those that lack D/G peaks clustered to the opposite side. As can be seen in both the plot, and the spectra, the darker samples typically feature more intense D/G bands. Worth noting are two dark samples that completely lack D/G peaks, placing them in the tight cluster that otherwise only contains light/white/colourless material. These two objects have, with all likelihood, been mischaracterized as quartzite. One of them is black with a fine texture, but also exhibits layering and resembles more a sandstone, possibly a sedimentary orthoquartzite. The second object likewise exhibits layering, but has a more rough texture with large quartz inclusions. 
PC 2 (`r raman.pc2var` %) captures similar features as PC 1, due to the main features being a presence or lack of D/G bands, which causes the "arch" visible in the plot. The "extremes" of PC 1 gravitate to the negative side of PC 2, with those that feature both distinct quartz and D/G bands clustering towards the positive side. PC 3 (`r raman.pc3var` %), however, captures no real features in the data and mainly describes the remaining noise left in the spectra after detrending and preprocessing the data.

@@

The analysis shows that ~14~Si makes up the main main bulk of the material, with ~13~Al recorded up to ~ 5% for some samples (fig. \@ref(fig:xrf-violin-plot)). Interestingly, the majority of the readings resulted in a ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown. This could both be due to the limits of the instrumentation, as well as the limitations in the sampling. Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content higher than 90%.

~12~Mg, ~23~V, ~25~Mn, ~39~Y and ~30~Zn recorded a high number missing values (below LOD), although in the case of ~12~Mg and ~25~Mn the values reached up to 0.3% and 3% respectively. Due to the number of missing values these elements were therefore excluded from the PCA. Similarly, as the analysis was carried out with field methods in mind, the objects were not cleaned as to remove potential remains from physical handling (e.g. P, S). Element variables that may be influenced by remains from physical management of the objects were thus also excluded. The following elements were included in the PCA:  ~13~Al, ~14~Si, ~19~K, ~20~Ca, ~22~Ti, ~26~Fe, ~38~Sr and ~40~Zr.

@@

The PCA resulted in 2 PCs with an eigenvalue above 1 (fig. \@ref(fig:xrf-pca-loadings-scree)), and a cumulative explained variance of `r xrf_var`%. K-means cluster analysis was then performed on the PCA data in order to help identify potential groups exhibiting similar chemofacies. Using the elbow graph method 3 clusters were deemed suitable for the analysis. As can be seen in the score plot (fig. \@ref(fig:xrf-pca-a)), the three K-means clusters remain separated. Along PC 1 cluster 3 is marked by a higher ~26~Fe and ~40~Zr content on average, as well as a ~14~Si content lower than 50%. It mainly features samples with a dark hue, with the few samples from the light group being grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue. Cluster 2 is separated out along PC 2, and includes only samples with a ~14~Si content above 90%, while ~26~Fe and ~40~Zr is low or below the LOD. The remaining elemental variables are all higher on average than the other clusters, as is to be expected considering the higher ~14~Si value. Cluster 1 includes the main bulk of the light group, as well as all of the *brecciated quartz* samples. As in the case of cluster 3, apart from one outlier, the remaining samples in cluster 1 have a ~14~Si content below 50%. While the clay mineral related elements (Al, Ca, K) are comparable to cluster 3, cluster 1 measured lower in ~22~Ti, ~26~Fe and ~40~Z. Where it stands out, however, is in the large amount of samples with ~38~Sr readings, as almost all ~38~Sr records come from samples in cluster 3. ~38~Sr is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays [@ahijado_etal05;@banner95].

@@

```{r xrf-pca-b, fig.cap= "xrf score plot", echo = FALSE}
#quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
knitr::include_graphics("../figures/009-xrf-pca-bNOSR.png", error = FALSE)
```

@@

Visualising the XRF k-means clusters onto the PCA for the NIRS data further emphasizes the variation within the general color groups (fig. \@ref(fig:nir-xrf-kmeans)). While the ~26~Fe/~40~Zr enriched samples largely are made up of those objects within the dark cluster in the NIR PCA. There are, however, a fair number of exceptions to this rule. A closer look at three of the objects that are part of the dark color group, but with different k-means, highlights the different nuances (fig. \@ref(fig:nir-sample-comparison)).

```{r nir-xrf-kmeans, fig.cap= "xrf boxplot by kmeans cluster", echo = FALSE}
knitr::include_graphics("../figures/010-nir-pca-xrf-kmeans.png", error = FALSE)
```

@@

```{r nir-sample-comparison, fig.cap= "xrf violin plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/011-sample-comparison.R"), echo = FALSE}
quietly(eval(parse("./analysis/paper/011-nir-sample-comparison.R", encoding = "UTF-8")))
knitr::include_graphics("../figures/011-nir-sample-comparison.png", error = FALSE)
```

@@

The most easily recognizable megascopic feature of petrographic material is its color, which, together with texture, has been the main attribute used for grouping quartzite in North Swedish archaeological research [e.g. @forsberg85 pp.60–64; @kaeck09 pp.53–55]. While it is true that some inference of elemental content can be made based on color this is significantly more difficult with heterogeneous material such as quartzite. Indeed, destructive techniques have been identified as necessary in order to document the metamorphic features that are necessary for distinguishing between sedimentary and metamorphic quartzite [@howard05; @prieto_etal19]. Non-destructive application of the above field-based spectroscopic techniques does, however, bring to light some interesting characteristics of the material collected in Vilhelmina and Åsele parish.

Despite the limitations inherent with dark and translucent (colourless) materials in the near-infrared when using the ASD, it does provide a good estimation of differences related to general hue. That being said, the results also highlights the complexity inherent with categorizing the color of a heterogeneous material which lacks a uniform surface. Although the surface area sampled by the contact probe is large enough to cover any potential variation in color it will ultimately generate an average spectra of these features. What differences could be observed between the samples in the spectra, apart from absorbance, were limited, and mainly occurred within the metal-OH range. This metal-OH overtone region could possibly relate to the presence of clay minerals (AlOH), such as muscovite, in the material.

The presence of the “graphite peaks” is another interesting aspect of the material revealed by the Raman analysis. These graphite-like carbon structures likely relate to the diagenetic environment in which the material formed, and should thus be critical to identifying the area of plausible geological origin. Interestingly, mentions of "graphitic quartzite" can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day. Considering the presence of Alum shale and similar carbon-rich tectonic strata within the Scandian mountain chain [@stephens_bergmanweihed20 pp. 482-ff], it would not be strange to encounter carbon structures in quartzite. Naturally, identifying the geological origin for these materials does not necessitate a direct link between that area and the procurement of the material, but it is nonetheless a key point for understanding the material flow within the region.

Sampling...xrf

>>>>

Petrographic characterization and provenance studies are instrumental in the modelling and inference of trade and exchange networks among prehistoric societies (REF). The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery. Petrographic characterization of the material represents the important first step in the reconstruction of the *chaîne opératoire*, as well as questions regarding raw material selection and management. Flint and obsidian have both been central to the study of trade and exhange networks in European prehistory, due in part to the well-surveyed geological sources for these. The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. The current bias that exists in the study of raw material management in regions where local flint and obsidian sources are non-existent^[This is not to say that flint was not available or used in these regions] has been highlighted by Prieto et al. [-@prieto_etal19 p. 15]. In Northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region. While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. While models of raw material sourcing and trade and exchange of slate and flint within Northern Sweden has been put forth [@halen94 pp. 115-122; @lundberg97 pp. 161-171], there has been little development on the topic of quartz and quartzite. A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the modern day landscape. The tentative interpretations of lithic material sourcing within Northern Sweden tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12]. This does not, however, account for the possibility that the material was sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250]. Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.

Prieto et al. [-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature, and different researchers have thus emphasized different features as being characteristic of the material [@howard05]. Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies. In the case of archaeological research it is worth considering situations where destructive techniques simply are not applicable. Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution. While destructive methods may prohibit access to such material there are still non-destructive alternatives that can generate valuable information. Aside from being non-destructive spectroscopic instrumentations feature fast measurements, making it an attractive option for field-based screening purposes. While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other instrumentations that may provide a deeper insight into the structural properties. The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. The same can be said for Raman spectroscopy, which has been applied extensively in the Cultural Heritage sector and in the analysis of artwork and manuscripts [@vandenabeele12].

@@

The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from excavated and surveyed sites, including stray finds, in Västerbotten County, Sweden. The sites are distributed along the Ångermanälven river valley, from the lake Kultsjön in the north, down towards the town of Åsele in the south. A dataset of `r length(tot_uniq_samples)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF. The material features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint, with some being almost fully translucent. The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent).

@@

The dataset includes three main groups of registered petrographic material: quartz, quartzite, and brecciated quartz (Swe. "*brecciekvarts*"). Those that had no prior classification available were sorted into either the quartz or quartzite category based on visual inspection. This resulted in: `r length(bquartz)` brecciated quartz samples, `r length(quartz)` quartz samples, and `r length(quartzite)` quartzite samples. The inclusion of quartz and brecciated quartz in the study serves two purposes. The first being to test the potential for separating quartz from quartzite based on their chemical and structural characteristic, and the second related purpose is to account for potential mistakes in the original classification.

When analysing the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67 p.8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden. Åhman describes the material as easily mistaken for an unusually pure quartzite, however, the fragmented configuration of quartz can be identified via microscopic analysis. *Brecciated quartz* in particular became a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g. @holm91 p.24]. As the material frequently has been characterized based on megascopic characteristics, despite the central features necessary for characterization being microscopic, there is value in including a test group in the study.

@@

Data analysis and visualisation was performed in R [@Rteam21] using a suite of packages, including Chemospec [@hanson16], missMDA [@josse_husson16], ggplot2 [@wickham16] and plotly [@sievert20]. Pre-processing and analysis of the data is different for each instrumentation and based on relevant literature. Data exploration and analysis was performed using Principal Component Analysis (PCA).

@@

NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe. The ASD features a detection range of 350 nm – 2 500 nm, with spectral sampling (resp. spectral resolution) of 1.4 nm (resp. 3 nm) in the visible and near infrared range and 1.1 nm (resp. 10 nm) in the short‐wave infrared range. Each object was measured against a white reference background and the collected measurements were averaged for each artefact. The data was restricted to the 1 000 nm - 2 500 nm range and mean-centred before PCA modelling.

@@

The XRF data was transformed using *centred log-ratio transformation*, mean-centered and scaled to unit variance before PCA. As the resulting dataset also features values below the Limit of Detection (LOD) these were treated as NULL (or "missing values") during PCA. A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement. Since missing values are not permitted during PCA these had to be imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data. Two methods for imputation can be implemented, iterative or regularized, and since the former suffers when there is a large number of missing values the latter was used.

Multivariate analysis methods, such as PCA, are also useful for overcoming some of the issues related to so-called "data closure" [@reimann_etal12]. The compositional data add up to a constant (100%), which is described as the *closed sum* effect. Correlation analysis is directly affected by this *closure* effect and therefore not suitable. The application of *centred log-ratio transformation* was done in order to address the closed sum effect, and in combination PCA it can reveal the structure and relations between the variables (elements) more accurately.

@@

\linebreak
<!--
## O2PLS

@@

O2PLS was implemented in R [@rcoreteam21] using the *OmicsPLS* package [@bouhaddani_etal18]. The data was pre-processed according to the above descriptions, and then cross-validation was used in order to determine the number of components to fit to the model. 
-->
\pagebreak

@@

Based on the evaluation of the spectra a couple of significant peaks can be identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm. These features are all the results of overtones and vibration combinations of hydroxyl (OH) groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water (fig. \@ref(fig:nir-savitzky-transform-alt) & \@ref(fig:nir-peak-summary)) [@hunt77 p. 508-510; @sciuto_etal19]. OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occuring in the 2 000 nm - 2 500 nm region. As a general rule, the 2 200 nm and 2 350 nm peak positions have been related to aluminium (Al) and magnesium (Mg) combinations [@hunt77 p.509; @clark_etal90 p.12664]. While there are exceptions, a more intense 2 200 nm band has typically been related to AlOH, and a more intense 2 350 nm band to MgOH. In the current spectra the 2 200 nm band is typically the more intense band.

\linebreak
 
```{r nir-savitzky-transform-alt, fig.cap= "SG transformed spectra contrasting the results between the spectra in the dark (black) and light (orange) groups", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform-alt.R"), echo = FALSE}
quietly(source("./analysis/paper/003-nir-savitzky-transform-alt.R", encoding = "UTF-8"))
knitr::include_graphics("../figures/003-nir-savitzky-transform-alt.png", error = FALSE)
```

@@

```{r nir-peak-summary, fig.cap= "Average NIR spectra for each colour group ", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/002-nir-peak-summary.R"), echo = FALSE}
quietly(source("./analysis/paper/002-nir-peak-summary.R", encoding = "UTF-8"))
knitr::include_graphics('../figures/002-nir-peak-summary.png', error = FALSE)
```

@@

While the explained variance of PC 3 is low, (`r nir.pc3var` %), it still contains a number of the aforementioned features visible in the spectra (inverse to PC 1) (fig. \@ref(fig:nir-pca-loadings)). Adding PC 3 to a score plot (fig. \@ref(fig:nir-pca-fig):B), however,  there seems to be little new information gained apart from a couple of samples, mainly from the light group, that break away from the remaining samples.

@@

The resulting PCA model (fig. \@ref(fig:raman-pca)) shows a clear trend along PC 1 (`r raman.pc1var` %) related to the quartz and D - G peaks. Samples where the D/G peaks dominate in terms of intensity are clustered towards the positive side of PC 1, with those that lack D/G peaks clustered to the opposite side. As can be seen in both the plot, and the spectra, the darker samples typically feature more intense D/G bands. Worth noting are two dark samples that completely lack D/G peaks, placing them in the tight cluster that otherwise only contains light/white/colourless material. These two objects have, with all likelihood, been mischaracterized as quartzite. One of them is black with a fine texture, but also exhibits layering and resembles more a sandstone, possibly a sedimentary orthoquartzite. The second object likewise exhibits layering, but has a more rough texture with large quartz inclusions. 
PC 2 (`r raman.pc2var` %) captures similar features as PC 1, due to the main features being a presence or lack of D/G bands, which causes the "arch" visible in the plot. The "extremes" of PC 1 gravitate to the negative side of PC 2, with those that feature both distinct quartz and D/G bands clustering towards the positive side. PC 3 (`r raman.pc3var` %), however, captures no real features in the data and mainly describes the remaining noise left in the spectra after detrending and pre-processing the data.

@@

The analysis shows that ~14~Si makes up the main main bulk of the material, with ~13~Al recorded up to ~ 5% for some samples (fig. \@ref(fig:xrf-violin-plot)). Interestingly, the majority of the readings resulted in a ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown. This could both be due to the limitations of the instrumentation, as well as the sampling. Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content higher than 90%.

~12~Mg, ~23~V, ~25~Mn, ~39~Y and ~30~Zn recorded a high number of missing values (below LOD), although in the case of ~12~Mg and ~25~Mn the values reached up to 0.3% and 3% respectively. Due to the number of missing values these elements were therefore excluded from the PCA. Similarly, as the analysis was carried out with field methods in mind, the objects were not cleaned as to remove potential remains from physical handling (e.g. P, S). Element variables that may be influenced by remains from physical management of the objects were thus also excluded. The following elements were included in the PCA:  ~13~Al, ~14~Si, ~19~K, ~20~Ca, ~22~Ti, ~26~Fe and ~40~Zr. The inclusion of ~38~Sr resulted in a couple of strong outliers and overall impacted the interpretation of the model, and was therefore excluded.

@@

The PCA resulted in 2 PCs with an eigenvalue above 1 (fig. \@ref(fig:xrf-pca-loadings-scree)), and a cumulative explained variance of `r xrf_var`%. K-means cluster analysis was then performed on the PCA data in order identify potential groups exhibiting similar chemofacies. Using the elbow graph method 3 clusters were deemed suitable for the analysis. As can be seen in the score plot (fig. \@ref(fig:xrf-pca-a)), the three K-means clusters remain separated. Cluster 1 is marked by a higher ~26~Fe and ~40~Zr content on average, as well as a ~14~Si content lower than 50%. It mainly features samples with a dark hue, with the few samples from the light group that are grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue. Cluster 3 is separated out along PC 2, and includes only samples with a ~14~Si content above 90%, while ~26~Fe and ~40~Zr is low or below the LOD. The remaining elemental variables are all higher on average than the other clusters, as is to be expected considering the higher ~14~Si value.

Cluster 2 includes the main bulk of the light group, as well as all of the *brecciated quartz* samples. As in the case of cluster 1, apart from a couple of outliers, most of the samples in cluster 2 have a recorded ~14~Si content below 50%. While the clay mineral related elements (Al, Ca, K) are comparable to cluster 1, cluster 2 measured lower in ~22~Ti, ~26~Fe and ~40~Z. Where it stands out, however, is in the large amount of samples with ~38~Sr readings, as almost all ~38~Sr records come from samples in cluster 2. ~38~Sr is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays [@ahijado_etal05;@banner95], which might be one explanation for its presence in cluster 2.

@@

@@

Visualising the K-means cluster analysis of the XRF PCA in the NIRS model further emphasizes the variation within the general color groups (fig. \@ref(fig:nir-xrf-kmeans)). Most of the dark material belongs to K-means cluster 1, which on average features higher ~22~Ti, ~26~Fe, and ~40~Zr values. This is not surprising as Fe may contribute to a darker hue in the material. Cluster 2, which features low Si values and clustered alongside clay mineral indicating elements, is fairly evenly distributed in the NIR PCA, including some of the darker objects. The same can be said about cluster 3, which indicates no real patterning in the NIR model.

This variation can be exemplified by looking closer at three samples selected from the dark dark group, that each belong to a different K-means cluster. They all share a similar hue with D/G peaks detected in the Raman analysis (fig. \@ref(fig:nir-sample-comparison)). All three samples have similar absorbance values, but sample 391 exhibits some differences in the spectra compared to the other two samples. Out of the three objects sample 391 seems to stand out more clearly from the others, with the XRF analysis being what mainly seems to separate the other two.

```{r nir-xrf-kmeans, fig.cap= "XRF PCA K-means cluster on NIR PCA", echo = FALSE}
knitr::include_graphics("../figures/010-nir-pca-xrf-kmeans.png", error = FALSE)
```

@@

```{r nir-sample-comparison, fig.cap= "sample comparison", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/011-sample-comparison.R"), echo = FALSE}
quietly(eval(parse("./analysis/paper/011-nir-sample-comparison.R", encoding = "UTF-8")))
knitr::include_graphics("../figures/011-nir-sample-comparison.png", error = FALSE)
```

@@

The study represents a good overview of the range of material which exists within the frames of what archaeologists refer to as *quartzite*, at least within Northern Sweden. Visually the material varies greatly, both in terms of texture as well as color. The initial classification of the material is primarily based on the archaeological documentation from the surveys and excavations, and as was to be expected there seems to be some possible mischaracterization due to this. The intention of keeping these original classifications was to further test whether the instrumentation could pick up on these possible discrepancies. These instances are not represented as outliers in the different PCA models, however, but rather distinguish themselves by behaving contrary to expectations. In a similar manner the *brecciated quartz* exhibited no unique signals in the analysis, and behaved much like the other semi-translucent material in the dataset. The two quartz samples, on the other hand, generated somewhat different results. While they did not contrast with the dataset as a whole they placed in different cluster within the models, with only one them recording > 90% ~14~Si. This, with a visual inspection of the material, suggests that at least one of them have been mislabeled.

The most easily recognizable megascopic feature of petrographic material is its color, which, together with texture, has been the main attribute used for grouping quartzite in North Swedish archaeological research [e.g. @forsberg85 pp.60–64; @kaeck09 pp.53–55]. While it is true that some inference of elemental content can be made based on color this is significantly more difficult with heterogeneous material such as quartzite. Indeed, destructive techniques have been identified as necessary in order to document the metamorphic features that are necessary for distinguishing between sedimentary and metamorphic quartzite [@prieto_etal19]. Alhough field-based megascopic methods for petrographic characterization of quartzite have been proposed [e.g. @howard05], it would be fair to suggest that these have seen little to no use among archaeologists. As such, archaeological classification of quartzite is mainly based on a general understanding of the material. Development of field-based spectroscopic methods is therefore desirable, as these not only would increase the accuracy, but also open up for comparative analysis of the material and aid in guiding the survey and excavation efforts.

The three instrumentations (NIRS, Raman, XRF) are capable of characterizing some aspects of quartzite. They highlight the complexity of the heterogeneous material, as well as the limitations of restricting oneself to one singular technique. Despite the limitations inherent with dark and translucent (colourless) materials in the near-infrared when using the ASD, it does provide a good estimation of differences related to general hue. That being said, the results also highlights the complexity inherent with categorizing the color of a heterogeneous material which lacks a uniform surface. Although the surface area sampled by the contact probe is large enough to cover any potential variation in color it will ultimately generate an average spectra of these features. Including other techniques, such as imaging techniques [@sciuto_etal19], or Vis-spectrophotometers [@ramacciotti_etal22], would open up for a more systematic way of characterizing color. Apart from the absorbance of the material, what differences could be observed between the samples in the NIR spectra were limited, and mainly occurred within the metal-OH range.
